---
title: "Analysis"
author: "Takashi Iida"
date: today
format: html
html-table-processing: none
---
## Overview

This document creates tables and figures for the manuscript.

## Setup

```{r}
#| label: setup
#| message: false


library(tidyverse)
library(haven)
library(gtsummary)
```

## Load clean data  

```{r}
# data import
# install.packages("haven")


getwd()

base_directory <- "/Users/iidatakashi/Documents/Git/HRS-E2-fellowship/data"

setwd(base_directory)

dat_long <- readRDS("dat_long.rds")
head(dat_long)

names(dat_long)
```


## Table 1

```{r}
#| label: tbl-table1
#| tbl-cap: "Table 1. Baseline characteristics at Wave 8 (2006), prior to multiple imputation. 
#|   Values are mean (SD) for continuous variables and n (%) for categorical variables.
#|   Missing data: income (XX%), smoking (XX%), comorbidity index (XX%)."
dat_long |>
  filter(wave == 8) |> 
  mutate(
    housing_stress_bin = factor(
      housing_stress_bin,
      levels = c(1, 0),
      labels = c("Housing Stress", "No Housing Stress")
    ),
    # sex -> factor
    sex = factor(
      sex,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ),
    education = factor(
      education,
      levels = c(1,2,3,4,5), 
      labels = c("Less than high school", "General Educational Diploma", "High school diplomas", "Some college", "College and above")
    ),
    # comorbidity_index, ser7 -> numeric
    comorbidity_index = as.numeric(comorbidity_index),
    ser7 = as.numeric(ser7)
  ) |>
  select(
    housing_stress_bin,
    age, sex, race_etn, education, income, smoking,
    comorbidity_index,
    imrc, dlrc, ser7
  ) |>
  tbl_summary(
    by = housing_stress_bin,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)",
      income            ~ "{median} ({p25}, {p75})" 
    ),
    type = list(
      comorbidity_index ~ "continuous",
      ser7              ~ "continuous"
    ),
    label = list(
      age               ~ "Age, mean(SD)(years)",
      sex               ~ "Sex, n(%)",
      race_etn          ~"Race and ethnicity, n(%)",
      education         ~ "Education, n(%)",
      income            ~ "Household income, median(IQR)($)",
      smoking           ~ "Current smoker, n(%)",
      comorbidity_index ~ "Comorbidity index (0–8), mean(SD)",
      imrc              ~ "Immediate word recall (0–10), mean(SD)",
      dlrc              ~ "Delayed word recall (0–10), mean(SD)",
      ser7              ~ "Serial 7's (0–5), mean(SD)"
    ),
    missing = "no"
  ) |>
  add_overall() |>
  modify_header(
    stat_1 ~ "**Housing Stress**  \n(n = {n})",
    stat_2 ~ "**No Housing Stress**  \n(n = {n})",
    stat_0 ~ "**Overall**  \n(n = {n})"
  ) |>
  bold_labels()
```






## Multiple imputation


```{r}
# install.packages("mice")
library(mice)

# check num of NAs in each variable
colSums(is.na(dat_long |> 
    select(housing_stress_bin, total_recall_z, ser7_z,
            age, income, smoking, comorbidity_index)))

            library(mice)

# dataset for multiple imutation
dat_mi <- dat_long |> 
  select(
    HHIDPN, wave,
    
    # variabels for analysis
    housing_stress_bin,
    total_recall_z, ser7_z,
    age, income, smoking, comorbidity_index,
    
    # Time-invariant
    sex, race_etn, education,
    
    # Auxiliary variables（attritionと相関しやすい）
    # imrc, dlrc,        # components of total_recall_z ->remove(collinear)
    # ser7,              # raw score ->remove(collinear)
    # total_recall,    # raw total recall ->remove(collinear)
    # housing_stress,    # raw housing stress ->remove(collinear)
  )

# remove HHIDPN and wave for imputation
ini <- mice(dat_mi, maxit = 0)
meth <- ini$method
meth["HHIDPN"] <- ""
meth["wave"]   <- ""

# Warning
# ini$loggedEvents
# table(dat_mi$race_etn, useNA = "always")

# remove HHIDPN from predictormatrix
ini  <- mice(dat_mi, maxit = 0)
meth <- ini$method
meth["HHIDPN"] <- ""
meth["wave"]   <- ""

# edit predictormatrix
pred <- ini$predictorMatrix
pred[, "HHIDPN"] <- 0  # remove HHIDPNfrom predictor

imp <- mice(
  dat_mi, 
  method          = meth,
  predictorMatrix = pred,
  m               = 20, 
  seed            = 123, 
  print           = FALSE
)

# check
summary(imp)
plot(imp)
```


## Modeling


```{r}
install.packages("fixest")
library(fixest)

# imputed datasetsを取り出してリストに格納
imp_list <- lapply(1:20, function(i) complete(imp, i))

# 各imputed datasetにFEモデルを適用
models_tr <- lapply(imp_list, function(d) {
  feols(total_recall_z ~ housing_stress_bin + age + income + 
                         smoking + comorbidity_index + wave 
        | HHIDPN,
        cluster = ~HHIDPN,
        data = d)
})

models_ser7 <- lapply(imp_list, function(d) {
  feols(ser7_z ~ housing_stress_bin + age + income + 
                 smoking + comorbidity_index + wave 
        | HHIDPN,
        cluster = ~HHIDPN,
        data = d)
})

# Rubin's rulesでpooling
pool_tr   <- pool(models_tr)
pool_ser7 <- pool(models_ser7)

summary(pool_tr, conf.int = TRUE)
summary(pool_ser7)
```


## Plot

### Forest plot

```{r}
ggplot(res_plot, aes(x = estimate, y = outcome, color = outcome)) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = `2.5 %`, xmax = `97.5 %`), 
                 height = 0.15, size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  # p値をラベルとして追加
  geom_text(aes(label = paste0("p = ", round(p.value, 3)),
                x = `97.5 %` + 0.005),
            hjust = 0, size = 3.5, color = "gray30") +
  scale_color_manual(values = c("Episodic Memory" = "#E07B6A",
                                "Working Memory"  = "#4AABB8")) +
  labs(
    title = "Effect of Ongoing Housing Stress\non Cognitive Function",
    subtitle = "Fixed Effects Model\nadjusted for age, income, smoking, comorbidity",
    x = "Coefficient (95% CI)",
    y = NULL
  ) +
  expand_limits(x = 0.02) +  # p値表示のスペース確保
  theme_minimal() +
  theme(legend.position = "none",
        plot.title    = element_text(size = 13, face = "bold"),
        plot.subtitle = element_text(size = 10, color = "gray50"),
        axis.text.y   = element_text(size = 12))
```



## DAG

Description: This DAG illustrates the hypothesized relationship between housing stress (exposure) and cognitive decline (outcome). Demographic characteristics, health problems, social determinants of health (SDoH), and smoking status are included as covariates. Demographic characteristics are age, sex, and race and ethnicity. SDoH are education and household incomes. Health problems include hypertension, diabetes or high blood sugar, cancer or malignant tumor (excluding skin cancer), chronic lung disease (excluding asthma), heart disease (including heart attack, coronary heart disease, angina, congestive heart failure, or other heart problems), stroke or transient ischemic attack, emotional, nervous, or psychiatric problems, and arthritis or rheumatism.


Abbreviations: DAG = Directed Acyclic Graph, SDoH = Social Determinants of Health


```{dot}
digraph DAG {
  layout=neato;
  node [shape=box, fontsize=12];
  
  Demographic     [label="Demographic",         pos="0,2!"];
  Health_Problems [label="Health Problems",     pos="2,3!"];
  SDoH            [label="SDoH",                pos="2,1!"];
  Health_Behavior [label="Smoking Status",     pos="2,0!"];
  Housing_Stress  [label="Housing Stress (X)",  pos="4,1.5!"];
  Cognitive_Decline [label="Cognitive Decline (Y)", pos="6,1.5!"];
  
  Demographic -> Health_Problems;
  Demographic -> SDoH;
  Demographic -> Housing_Stress;
  Demographic -> Cognitive_Decline;
  
  Health_Problems -> SDoH;
  Health_Problems -> Housing_Stress;
  Health_Problems -> Cognitive_Decline;
  
  SDoH -> Housing_Stress;
  SDoH -> Cognitive_Decline;
  
  Health_Behavior -> Housing_Stress;
  Health_Behavior -> Cognitive_Decline;
  
  Housing_Stress -> Cognitive_Decline;
}
```