---
title: "Analysis"
author: "Takashi Iida"
date: today
format: html
html-table-processing: none
---
## Overview

This document creates tables and figures for the manuscript.

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(haven)
```

## Load clean data 

- raw data: dat
- analysis data: dat_an

```{r}
# data import
# install.packages("haven")


getwd()

base_directory <- "/Users/iidatakashi/Documents/Git/HRS-E2-fellowship/data"

setwd(base_directory)

dat <- read_sas("randhrs1992_2022v1.sas7bdat")

```

## Overall data exploration

```{r}
#| label: overall-exploration
#| echo: true
#| message: false


head(dat) 
dim(dat)  # 45234 19880: n. of observation and n. of variables
names(dat) # names of variables
```

## Variables names
Variable structure
- [Person/Unit][Wave][Topic...]
    - 1st character: who the variable refers to
        - R = respondent (reference person)
        - S = spouse/partner
        - H = household
    - 2nd character: wave identifier
        - 1,2,3,... = specific wave
        - sometimes A = “all waves” / not wave-specific (varies by item)

person id: HHIDPN (HHIDPN: HHold ID + Person Number /Num)


```{r}


dat %>% group_by("HHIDPN") |> select(contains("HOUSEPRB")) 
nrow(dat %>% group_by("HHIDPN"))

```



## Exposure exploration

- Exposure is Ongoing housing problems. 
- Identifier is HOUSEPRB in the colmun names. 
- Includde waves from 8 to 13 (10 years)

```{r}
#| label: exposure-exploration
#| echo: false

# numbers of waves containing HOUSEPRB
dim(dat %>% select(contains("HOUSEPRB"))) #  16

# number of participants for HOUSEPRB at wave 8-15
## wave 9 did not have HOUSEPRB
dat |>
  summarise(
    w8_housing = sum(!is.na(R8LBHOUSEPRB)), # 7422
    w10_housing = sum(!is.na(R10LBHOUSEPRB)), # 5319
    w11_housing = sum(!is.na(R11LBHOUSEPRB)), # 1 (no f/u subgrp )
    w12_housing = sum(!is.na(R12LBHOUSEPRB)), # 4450
    w13_housing = sum(!is.na(R13LBHOUSEPRB)), # 3
    w14_housing = sum(!is.na(R14LBHOUSEPRB)), # 2800
    w15_housing = sum(!is.na(R15LBHOUSEPRB)) # 2800
  )

# waves: S&R 8,10,11,12,13,14,15,16
names(dat %>% select(contains("HOUSEPRB"))) 

# select person_id,  House problem from w8 to w13
# make lists 
list_id_husplb <- c("HHIDPN", "R8LBHOUSEPRB", "R10LBHOUSEPRB","R11LBHOUSEPRB", "R12LBHOUSEPRB", "R13LBHOUSEPRB", "R14LBHOUSEPRB" )
# names(dat |> select(list_id_husplb, list_IMRC))

# Spouse lists
list_id_husplb_spo <- c("HHIDPN", "S8LBHOUSEPRB", "S10LBHOUSEPRB","S11LBHOUSEPRB", "S12LBHOUSEPRB", "S13LBHOUSEPRB", "S14LBHOUSEPRB" )
```


## Outcome-exploration

- Immediate Word Recall: IMRC
    - memo: RwFIMRC, RwFIMRCP, and RwFIMRCW indicate whether the value was imputed (=1) or not (=0). -> we will not use
- Delayed Word Recall
    - memo: RwFDLRC, RwFDLRCP, and RwFDLRCW indicate whether the value was imputed (=1) or not (=0). 
- Serial 7’s
- Wave 9 (2008) was excluded from the analysis as housing stress data were not collected in that wave

```{r}

#| label: outcome-exploration
#| echo: false

# Immediate word recall
names(dat |> select(contains("IMRC")))
list_IMRC <- c("R8IMRC","R10IMRC","R11IMRC", "R12IMRC", "R13IMRC", "R14IMRC")
list_IMRC_spo <- c("S8IMRC","S10IMRC","S11IMRC", "S12IMRC", "S13IMRC", "S14IMRC")

# Delayed word recall
names(dat |> select(contains("DLRC")))
list_DLRC <- c("R8DLRC","R10DLRC","R11DLRC", "R12DLRC", "R13DLRC", "R14DLRC")
list_DLRC_spo <- c("S8DLRC","S10DLRC","S11DLRC", "S12DLRC", "S13DLRC", "S14DLRC")

# Series 7s
names(dat |> select(contains("SER7")))
list_SER7 <- c("R8SER7","R10SER7","R11SER7", "R12SER7", "R13SER7", "R14SER7")
list_SER7_spo <- c("S8SER7","S10SER7","S11SER7", "S12SER7", "S13SER7", "S14SER7")

# Cognitive score 27 (for sensitivity analysis)
names(dat |> select(contains("COG27")))
list_cog27 <- c("R8COG27","R10COG27","R11COG27", "R12COG27", "R13COG27", "R14COG27")
list_cog27_spo <- c("S8COG27","S10COG27","S11COG27", "S12COG27", "S13COG27", "S14COG27")

```



## Confounding exploration
variables to exlude 
- memory-related disease
    - whether or not the Respondent reports a doctor has ever told
them that they had this condition in this or any previous wave
    - 0: No
    - 1: Yes
- memo
    - dementia, alzheimer disease were measured from wave 10


variabels in table 1
- age
    - age at the begging of the interview in years
- sex
    - 1: Male
    - 2: Female
- Race
    - 1.White/Caucasian 
    - 2.Black/African American 
    - 3.Other
- Ethnicity
    - 0.Not Hispanic
    - 1.Hispanic
- education
    - 1. Lt High-school
    - 2. GED, 
    - 3. High-school graduate
    - 4. some college
    - 5. college and above
    - 1. Less than high school
    - 2. high school and general educational diplomas
    - 3. some College
    - 4. ollege and above
- income
    - total household income
- smoking
    - current smoking status
    - 0: No
    - 1: Yes
- Doctor diagnosied health problems: ever have condition
    - high BP
    - diabetes
    - cancer
    - lung disease
    - heart problem
    - stroke
    - psychiatic problem
    - arthritis
    - answer
        - 0: No
        - 1: Yes
    - memo
        - sleep disorder is measured from w13, so not included


```{r}
# exclusion variabels

## dementia ever
names(dat |> select(contains("MEMRYE")))
list_exclude <- c("R8MEMRYE")
list_exclude_spo <- c("S8MEMRYE")

dat |> select(list_exclude) |> filter(R8MEMRYE==1) |> nrow() # 978
dat |> select(list_exclude_spo) |> filter(S8MEMRYE==1) |> nrow() # 431



# variabels in table 1

## age at the begging of the interview in years
names(dat |> select(contains("AGEY"))) # R8AGEY_B
names(dat |> select(contains("AGEY"))) # S8AGEY_B

## Gender of respondents
## 1: Male
## 2: Female
names(dat |> select(contains("RAGENDER"))) # RAGENDER

## Race
## 1.White/Caucasian 
## 2.Black/African American 
## 3.Other
names(dat |> select(contains("RARACEM"))) # RARACEM


## Ethnicity
## 0.Not Hispanic
## 1.Hispanic
names(dat |> select(contains("RAHISPAN"))) # RAHISPAN

## Education
## 1. Lt High-school
## 2. GED, 
## 3. High-school graduate
## 4. some college
## 5. college and above
# change into
## 1. Less than high school
## 2. high school and general educational diplomas
## 3. some College
## 4. ollege and above
names(dat |> select(contains("RAEDUC"))) # RAEDUC


## Income:
## total household income
## cont : nominal dollars
## from w8 to w13
# "H8ITOT"  "H9ITOT"  "H10ITOT" "H11ITOT"
# "H12ITOT" "H13ITOT"
names(dat |> select(contains("ITOT"))) 

## Smoking
## 0: No
## 1: Yes
names(dat |> select(contains("SMOKEN"))) # R8SMOKEN, S8SMOKEN

list_confound <- c(
    "R8AGEY_B", "R10AGEY_B",   # age
    "R11AGEY_B", "R12AGEY_B", "R13AGEY_B",
    "RAGENDER", # gender
    "RARACEM",  # race
    "RAHISPAN", # ethnicity
    "RAEDUC", # education
    "H8ITOT",  "H10ITOT", # income
    "H11ITOT","H12ITOT", "H13ITOT",
    "R8SMOKEN", "R10SMOKEN", "R11SMOKEN", # smoking
    "R12SMOKEN", "R13SMOKEN"
    )

list_confound_spo <- c(
    "S8AGEY_B", "S10AGEY_B",   # age
    "S11AGEY_B", "S12AGEY_B", "S13AGEY_B",
    "RAGENDER", # gender (time-invariant, same variable)
    "RARACEM",  # race (time-invariant, same variable)
    "RAHISPAN", # ethnicity (time-invariant, same variable)
    "RAEDUC", # education (time-invariant, same variable)
    "H8ITOT",  "H10ITOT", # income (household-level, same variable)
    "H11ITOT","H12ITOT", "H13ITOT",
    "S8SMOKEN", "S10SMOKEN", "S11SMOKEN", # smoking
    "S12SMOKEN", "S13SMOKEN"
    )


## Doctor diagnosied health problems
### High BP
names(dat |> select(contains("HIBP")))
list_BP_1     <- c("R8HIBPE", "R10HIBPE","R11HIBPE", "R12HIBPE","R13HIBPE")
list_BP_1_spo <- c("S8HIBPE", "S10HIBPE","S11HIBPE", "S12HIBPE","S13HIBPE")

### Diabetes
names(dat |> select(contains("DIABE")))
list_db_2     <- c("R8DIABE", "R10DIABE", "R11DIABE", "R12DIABE", "R13DIABE")
list_db_2_spo <- c("S8DIABE", "S10DIABE", "S11DIABE", "S12DIABE", "S13DIABE")

### Cancer
names(dat |> select(contains("CANCRE")))
list_cancr_3     <- c("R8CANCRE", "R10CANCRE", "R11CANCRE", "R12CANCRE", "R13CANCRE")
list_cancr_3_spo <- c("S8CANCRE", "S10CANCRE", "S11CANCRE", "S12CANCRE", "S13CANCRE")

### Lung disease
names(dat |> select(contains("LUNGE")))
list_lung_4     <- c("R8LUNGE", "R10LUNGE", "R11LUNGE", "R12LUNGE","R13LUNGE")
list_lung_4_spo <- c("S8LUNGE", "S10LUNGE", "S11LUNGE", "S12LUNGE","S13LUNGE")

### Heart problem
names(dat |> select(contains("HEARTE")))
list_heart_5     <- c("R8HEARTE", "R10HEARTE", "R11HEARTE", "R12HEARTE", "R13HEARTE")
list_heart_5_spo <- c("S8HEARTE", "S10HEARTE", "S11HEARTE", "S12HEARTE", "S13HEARTE")

### Stroke
names(dat |> select(contains("STROKE")))
list_stroke_6     <- c("R8STROKE", "R10STROKE", "R11STROKE", "R12STROKE", "R13STROKE")
list_stroke_6_spo <- c("S8STROKE", "S10STROKE", "S11STROKE", "S12STROKE", "S13STROKE")

### Psychiatric Problems
names(dat |> select(contains("PSYCHE")))
list_psych_7     <- c("R8PSYCHE", "R10PSYCHE", "R11PSYCHE", "R12PSYCHE", "R13PSYCHE")
list_psych_7_spo <- c("S8PSYCHE", "S10PSYCHE", "S11PSYCHE", "S12PSYCHE", "S13PSYCHE")

### Arthritis
names(dat |> select(contains("ARTHRE")))
list_arthr_8     <- c("R8ARTHRE", "R10ARTHRE", "R11ARTHRE", "R12ARTHRE", "R13ARTHRE")
list_arthr_8_spo <- c("S8ARTHRE", "S10ARTHRE", "S11ARTHRE", "S12ARTHRE", "S13ARTHRE")


## list Comorbidity index
list_Comorbidity_index <- c(list_BP_1,
  list_db_2, list_cancr_3, list_lung_4,
  list_heart_5, list_stroke_6, list_psych_7,
  list_arthr_8)

list_Comorbidity_index_spo <- c(list_BP_1_spo,
  list_db_2_spo, list_cancr_3_spo, list_lung_4_spo,
  list_heart_5_spo, list_stroke_6_spo, list_psych_7_spo,
  list_arthr_8_spo)                                  
                   

# try
#dat |> select(contains("H8ITOT")) |> head()
#dat |> select(contains("RAEDUC")) |> unique()
#dat |> select(list_Comorbidity_index) 

```



## Create person-wave long data 

```{r}

waves_all <- c(8, 10, 12, 14)
treatment_waves <- c(8, 10, 12)

# Function to classify interview and vital status
classify_iwstat <- function(x) {
  dplyr::case_when(
    x == 1         ~ "alive_interviewed",
    x == 4         ~ "alive_nonrespondent",
    x %in% c(5, 6) ~ "dead",
    x == 7         ~ "dropped",
    x == 0         ~ "inapplicable",
    x == 9         ~ "unknown_vital_status",
    TRUE           ~ NA_character_
  )
}

person_long_all <- purrr::map_dfr(waves_all, function(w) {

  dat |>
    transmute(
      # Person ID
      HHIDPN,

      # Fixed baseline household ID for clustering
      cluster_id = H8HHID,

      # Household ID at each wave
      hhid_wave = .data[[paste0("H", w, "HHID")]],

      wave = w,

      # Baseline eligibility variable
      baseline_memory_disease = R8MEMRYE,

      # Time-invariant variables
      sex       = RAGENDER,
      race      = RARACEM,
      hispanic  = RAHISPAN,
      education = RAEDUC,

      # Exposure
      housing_stress =
        .data[[paste0("R", w, "LBHOUSEPRB")]],


      # Outcomes
      imrc = if (w < 14) {
        .data[[paste0("R", w, "IMRC")]]
      } else {
        coalesce(
          .data[[paste0("R", w, "IMRCP")]],
          .data[[paste0("R", w, "IMRCW")]]
        )
      },

      dlrc = if (w < 14) {
        .data[[paste0("R", w, "DLRC")]]
      } else {
        coalesce(
          .data[[paste0("R", w, "DLRCP")]],
          .data[[paste0("R", w, "DLRCW")]]
        )
      },

      ser7 = if (w < 14) {
        .data[[paste0("R", w, "SER7")]]
      } else {
        coalesce(
          .data[[paste0("R", w, "SER7P")]],
          .data[[paste0("R", w, "SER7W")]]
        )
      },

      # COG27(original)
      cog27 = .data[[paste0("R", w, "COG27")]],

      # Cognition measurement mode
      cognition_mode = if (w < 14) {
        case_when(
          !is.na(.data[[paste0("R", w, "IMRC")]]) |
            !is.na(.data[[paste0("R", w, "DLRC")]]) |
            !is.na(.data[[paste0("R", w, "SER7")]]) ~ "FTF/TEL",
          TRUE ~ NA_character_
        )
      } else {
        case_when(
          !is.na(.data[[paste0("R", w, "IMRCP")]]) |
            !is.na(.data[[paste0("R", w, "DLRCP")]]) |
            !is.na(.data[[paste0("R", w, "SER7P")]]) ~ "FTF/TEL",

          !is.na(.data[[paste0("R", w, "IMRCW")]]) |
            !is.na(.data[[paste0("R", w, "DLRCW")]]) |
            !is.na(.data[[paste0("R", w, "SER7W")]]) ~ "Web",

          TRUE ~ NA_character_
        )
      },

      # Time-varying confounders
      age =
        .data[[paste0("R", w, "AGEY_B")]],

      income =
        .data[[paste0("H", w, "ITOT")]],

      smoking =
        .data[[paste0("R", w, "SMOKEN")]],

      hibp =
        .data[[paste0("R", w, "HIBPE")]],

      diabetes =
        .data[[paste0("R", w, "DIABE")]],

      cancer =
        .data[[paste0("R", w, "CANCRE")]],

      lung =
        .data[[paste0("R", w, "LUNGE")]],

      heart =
        .data[[paste0("R", w, "HEARTE")]],

      stroke =
        .data[[paste0("R", w, "STROKE")]],

      psychiatric =
        .data[[paste0("R", w, "PSYCHE")]],

      arthritis =
        .data[[paste0("R", w, "ARTHRE")]],

      # Raw interview/vital status
      iwstat_raw =
        .data[[paste0("R", w, "IWSTAT")]]
    )
}) |>
  arrange(HHIDPN, wave)
```



## Create derived variables

```{r}

# comorbidity index
comorb_vars <- c(
  "hibp",
  "diabetes",
  "cancer",
  "lung",
  "heart",
  "stroke",
  "psychiatric",
  "arthritis"
)

person_long_all <- person_long_all |>
  mutate(
    # Total episodic memory score
    total_recall = imrc + dlrc,

    # Individual-level housing stress
    housing_stress_bin_ind = case_when(
      housing_stress == 1     ~ 0L,
      housing_stress %in% 2:4 ~ 1L,
      is.na(housing_stress)   ~ NA_integer_,
      TRUE                    ~ NA_integer_
    ),

    # Number of missing comorbidity items
    n_comorb_missing =
      rowSums(is.na(pick(all_of(comorb_vars)))),

    # Comorbidity index: defined only if all 8 items are observed
    comorbidity_index = if_else(
      n_comorb_missing == 0,
      rowSums(pick(all_of(comorb_vars))),
      NA_real_
    ),

    # Vital/interview status
    status = classify_iwstat(iwstat_raw),

    # Binary cognitive impairment
    cognitive_impairment = case_when(
      is.na(cog27) ~ NA_integer_,
      cog27 <= 11  ~ 1L,
      cog27 >= 12  ~ 0L
    ),

    treatment_wave = wave %in% treatment_waves
  )
```



## Create household-level exposure 

- If one of the members in one household, the household-level exposure : 1

```{r}

# create household-level exposure
household_exposure <- person_long_all |>
  filter(
    wave %in% treatment_waves,
    !is.na(hhid_wave)
  ) |>
  group_by(wave, hhid_wave) |>
  summarise(
    n_household_members =
      n_distinct(HHIDPN),

    n_housing_reports =
      sum(!is.na(housing_stress_bin_ind)),

    housing_stress_bin_hh = case_when(
      # No housing-stress report in the household
      n_housing_reports == 0 ~ NA_integer_,

      # At least one member reports housing stress
      any(
        housing_stress_bin_ind == 1L,
        na.rm = TRUE
      ) ~ 1L,

      # At least one report, and all observed reports are negative
      TRUE ~ 0L
    ),

    .groups = "drop"
  )
```

## Define baseline analytic cohort 

```{r}

eligible_ids <- person_long_all |>
  filter(
    wave == 8,
    baseline_memory_disease == 0,
    !is.na(housing_stress_bin_ind),
    !is.na(imrc),
    !is.na(dlrc),
    !is.na(ser7)
  ) |>
  distinct(HHIDPN) |>
  pull(HHIDPN)

# join household_exposure
dat_long_combined <- person_long_all |>
  filter(HHIDPN %in% eligible_ids) |>
  left_join(
    household_exposure,
    by = c("wave", "hhid_wave")
  ) |>
  arrange(HHIDPN, wave)

```


## Recode baseline variables

- factor
  - sex, race, education

```{r}

dat_long_combined <- dat_long_combined |>
  mutate(
    sex = factor(
      sex,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ),

    race_ethnicity = factor(
      case_when(
        hispanic == 1 ~ "Hispanic",
        race == 1     ~ "Non-Hispanic White",
        race == 2     ~ "Non-Hispanic Black",
        race == 3     ~ "Non-Hispanic Other",
        TRUE          ~ NA_character_
      ),
      levels = c(
        "Non-Hispanic White",
        "Non-Hispanic Black",
        "Hispanic",
        "Non-Hispanic Other"
      )
    ),

    education_cat = factor(
      case_when(
        education == 1         ~ 1L,
        education %in% c(2, 3) ~ 2L,
        education == 4         ~ 3L,
        education == 5         ~ 4L,
        TRUE                   ~ NA_integer_
      ),
      levels = 1:4,
      labels = c(
        "Less than high school",
        "High school or GED",
        "Some college",
        "College and above"
      )
    )
  )
```


# create z-scores

- scale(): (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)

```{r}
dat_long_combined <- dat_long_combined |>
  group_by(wave) |>
  mutate(
    total_recall_z =
      as.numeric(scale(total_recall)),

    ser7_z =
      as.numeric(scale(ser7))
  ) |>
  ungroup()

# check distribution
dat_long_combined |>
  group_by(wave) |>
  summarise(
    mean_total_recall_z = mean(total_recall_z, na.rm = TRUE),
    sd_total_recall_z   = sd(total_recall_z, na.rm = TRUE),
    mean_ser7_z         = mean(ser7_z, na.rm = TRUE),
    sd_ser7_z           = sd(ser7_z, na.rm = TRUE),
    .groups = "drop"
  )

dat_long_combined |>
  group_by(wave) |>
  summarise(
    n_total_recall     = sum(!is.na(total_recall)),
    n_total_recall_z   = sum(!is.na(total_recall_z)),
    n_ser7             = sum(!is.na(ser7)),
    n_ser7_z           = sum(!is.na(ser7_z)),
    .groups = "drop"
  ) 
```


## Checks 


```{r}

# One row per person per wave
stopifnot(
  !anyDuplicated(
    dat_long_combined[c("HHIDPN", "wave")]
  )
)

# Every participant should have four rows
dat_long_combined |>
  count(HHIDPN) |>
  count(n)

stopifnot(
  nrow(dat_long_combined) ==
    n_distinct(dat_long_combined$HHIDPN) * 4
)

# Baseline status
dat_long_combined |>
  filter(wave == 8) |>
  count(status)

# Status by wave
dat_long_combined |>
  count(wave, status)

# Household exposure must be identical within household-wave
dat_long_combined |>
  filter(
    treatment_wave,
    !is.na(hhid_wave)
  ) |>
  group_by(wave, hhid_wave) |>
  summarise(
    n_exposure_values =
      n_distinct(
        housing_stress_bin_hh,
        na.rm = TRUE
      ),
    .groups = "drop"
  ) |>
  filter(n_exposure_values > 1)

# num of cupples
dat_long_combined |>
  filter(wave == 8) |>
  count(cluster_id, name = "n_people") |>
  count(n_people)


# chcke NA in cluster_id: 0
dat_long_combined |>
  filter(wave == 8) |>
  summarise(
    n = n(),
    n_missing_cluster_id = sum(is.na(cluster_id)),
    n_missing_hhid_wave  = sum(is.na(hhid_wave)),
    n_clusters = n_distinct(cluster_id, na.rm = TRUE)
  )

```

## check flowchart

```{r}

# One row per person at Wave 8
baseline_flow <- person_long_all |>
  filter(wave == 8) |>
  distinct(HHIDPN, .keep_all = TRUE)

# Step 0: all RAND HRS participants
flow_0 <- dat |>
  distinct(HHIDPN)

# Step 1: interviewed at Wave 8
flow_1 <- baseline_flow |>
  filter(iwstat_raw == 1)

# Step 2: no doctor-diagnosed memory-related disease at Wave 8
flow_2 <- flow_1 |>
  filter(baseline_memory_disease == 0)

# Step 3: baseline housing-stress data observed
flow_3 <- flow_2 |>
  filter(!is.na(housing_stress_bin_ind))

# Step 4: all baseline primary cognitive outcomes observed
flow_4 <- flow_3 |>
  filter(
    !is.na(imrc),
    !is.na(dlrc),
    !is.na(ser7)
  )

flow_counts <- tibble(
  stage = c(
    "All participants",
    "Wave 8 respondents",
    "No baseline memory-related disease",
    "Baseline housing-stress data observed",
    "Final analytic sample"
  ),
  n_remaining = c(
    n_distinct(flow_0$HHIDPN),
    n_distinct(flow_1$HHIDPN),
    n_distinct(flow_2$HHIDPN),
    n_distinct(flow_3$HHIDPN),
    n_distinct(flow_4$HHIDPN)
  )
) |>
  mutate(
    n_excluded = lag(n_remaining) - n_remaining
  )

flow_counts
```

## Save 

```{r}

saveRDS(
  dat_long_combined,
  "dat_long_combined.rds"
)

```
