---
title: "Analysis"
author: "Takashi Iida"
date: today
format: html
html-table-processing: none
---
## Overview

This document creates tables and figures for the manuscript.

## Setup

```{r}
#| label: setup
#| message: false


library(tidyverse)
library(haven)
```

## Load clean data 

- raw data: dat
- analysis data: dat_an

```{r}
# data import
# install.packages("haven")


getwd()

base_directory <- "/Users/iidatakashi/Documents/Git/HRS-E2-fellowship/data"

setwd(base_directory)

dat <- read_sas("randhrs1992_2022v1.sas7bdat")

```

## Overall data exploration

```{r}
#| label: overall-exploration
#| echo: true
#| message: false


head(dat)
dim(dat)  # n. of observation and n. of variables
names(dat) # names of variables
```

## Variables names
Variable structure
- [Person/Unit][Wave][Topic...]
    - 1st character: who the variable refers to
        - R = respondent (reference person)
        - S = spouse/partner
        - H = household
    - 2nd character: wave identifier
        - 1,2,3,... = specific wave
        - sometimes A = “all waves” / not wave-specific (varies by item)

person id: HHIDPN (HHIDPN: HHold ID + Person Number /Num)


```{r}


dat %>% group_by("HHIDPN") |> select(contains("HOUSEPRB")) 
nrow(dat %>% group_by("HHIDPN"))
nrow

```



## Exposure exploration

- Exposure is Ongoing housing problems. 
- Identifier is HOUSEPRB in the colmun names. 
- Includde waves from 8 to 13 (10 years)

```{r}
#| label: exposure-exploration
#| echo: false

# numbers of waves containing HOUSEPRB
dim(dat %>% select(contains("HOUSEPRB"))) #  16

# waves: S&R 8,10,11,12,13,14,15,16
names(dat %>% select(contains("HOUSEPRB"))) 

# select person_id,  House problem from w8 to w13
# make lists 
list_id_husplb <- c("HHIDPN", "R8LBHOUSEPRB", "R10LBHOUSEPRB","R11LBHOUSEPRB", "R12LBHOUSEPRB", "R13LBHOUSEPRB", "R14LBHOUSEPRB" )
names(dat |> select(list_id_husplb, list_IMRC))

```


# delete later
```{r}

view(dat |> filter(dat$HHIDPN == 10001010) |> select(contains("HOUSEPRB")))

table(dat$R16LBHOUSEPRB, useNA = "ifany") 
table(dat$S16LBHOUSEPRB, useNA = "ifany")# one example among several 

## Ongoing housing preoblems: Categorical variables
table(dat$R12LBHOUSEPRB, useNA = "ifany") 
table(dat$S12LBHOUSEPRB, useNA = "ifany")# one example among several waves
prop.table(table(dat$R12LBHOUSEPRB))

# check cols for Ongoing housing preoblems
dat %>% select(contains("HOUSEPRB"))
dim(dat %>% select(contains("HOUSEPRB")))
names(dat %>% select(contains("HOUSEPRB")))

# Number of completed interviews per wave
colSums(!is.na(dat %>% select(contains("HOUSEPRB"))))

# Total number of completed interviews (all waves combined)
sum(dat %>% select(contains("HOUSEPRB")), na.rm = TRUE)

# integrate into vector
all_house_values <- unlist(dat %>% select(contains("HOUSEPRB")))

# distribution
table(all_house_values, useNA = "ifany")

# proportion
prop.table(table(all_house_values))
```


## Outcome-exploration

- Immediate Word Recall: IMRC
    - memo: RwFIMRC, RwFIMRCP, and RwFIMRCW indicate whether the value was imputed (=1) or not (=0). -> we will not use
- Delayed Word Recall
    - memo: RwFDLRC, RwFDLRCP, and RwFDLRCW indicate whether the value was imputed (=1) or not (=0). 
- Serial 7’s
- Wave 9 (2008) was excluded from the analysis as housing stress data were not collected in that wave

```{r}
#| label: outcome-exploration
#| echo: false

# Immediate word recall
names(dat |> select(contains("IMRC")))
list_IMRC <- c("R8IMRC","R10IMRC","R11IMRC", "R12IMRC", "R13IMRC")

# Delayed word recall
names(dat |> select(contains("DLRC")))
list_DLRC <- c("R8DLRC","R10DLRC","R11DLRC", "R12DLRC", "R13DLRC")

# Series 7s
names(dat |> select(contains("SER7")))
list_SER7 <- c("R8SER7","R10SER7","R11SER7", "R12SER7", "R13SER7")
```


## Confounding exploration
variables to exlude 
- memory-related disease
    - whether or not the Respondent reports a doctor has ever told
them that they had this condition in this or any previous wave
    - 0: No
    - 1: Yes
- memo
    - dementia, alzheimer disease were measured from wave 10


variabels in table 1
- age
    - age at the begging of the interview in years
- sex
    - 1: Male
    - 2: Female
- Race
    - 1.White/Caucasian 
    - 2.Black/African American 
    - 3.Other
- Ethnicity
    - 0.Not Hispanic
    - 1.Hispanic
- education
    - 1. Lt High-school
    - 2. GED, 
    - 3. High-school graduate
    - 4. some college
    - 5. college and above
- income
    - total household income
- smoking
    - current smoking status
    - 0: No
    - 1: Yes
- Doctor diagnosied health problems: ever have condition
    - high BP
    - diabetes
    - cancer
    - lung disease
    - heart problem
    - stroke
    - psychiatic problem
    - arthritis
    - answer
        - 0: No
        - 1: Yes
    - memo
        - sleep disorder is measured from w13, so not included

Sex, Education were not included in adjusted model because of the feature of the model

```{r}
# exclusion variabels

## dementia ever
names(dat |> select(contains("MEMRYE")))
list_exclude <- c("R8MEMRYE")

dat |> select(list_exclude) |> filter(R8MEMRYE==1) |> nrow()


# variabels in table 1

## age at the begging of the interview in years
names(dat |> select(contains("AGEY"))) # R8AGEY_B

## Gender of respondents
## 1: Male
## 2: Female
names(dat |> select(contains("RAGENDER"))) # RAGENDER

## Race
## 1.White/Caucasian 
## 2.Black/African American 
## 3.Other
names(dat |> select(contains("RARACEM"))) # RARACEM


## Ethnicity
## 0.Not Hispanic
## 1.Hispanic
names(dat |> select(contains("RAHISPAN"))) # RAHISPAN

## Education
## 1. Lt High-school
## 2. GED, 
## 3. High-school graduate
## 4. some college
## 5. college and above
names(dat |> select(contains("RAEDUC"))) # RAEDUC


## Income:
## total household income
## cont : nominal dollars
## from w8 to w13
# "H8ITOT"  "H9ITOT"  "H10ITOT" "H11ITOT"
# "H12ITOT" "H13ITOT"
names(dat |> select(contains("ITOT"))) 

## Smoking
## 0: No
## 1: Yes
names(dat |> select(contains("SMOKEN"))) 

list_confound <- c(
    "R8AGEY_B", "R10AGEY_B",   # age
    "R11AGEY_B", "R12AGEY_B", "R13AGEY_B",
    "RAGENDER", # gender
    "RARACEM",  # race
    "RAHISPAN", # ethnicity
    "RAEDUC", # education
    "H8ITOT",  "H10ITOT", # income
    "H11ITOT","H12ITOT", "H13ITOT",
    "R8SMOKEN", "R10SMOKEN", "R11SMOKEN", # smoking
    "R12SMOKEN", "R13SMOKEN"
    )


## Doctor diagnosied health problems
### High BP
names(dat |> select(contains("HIBP"))) 
list_BP_1 <- c("R8HIBPE", "R10HIBPE","R11HIBPE", "R12HIBPE","R13HIBPE")

### Diabetes
names(dat |> select(contains("DIABE"))) 
list_db_2 <- c("R8DIABE", "R10DIABE", "R11DIABE", "R12DIABE", "R13DIABE")

### Cancer
names(dat |> select(contains("CANCRE"))) 
list_cancr_3 <- c("R8CANCRE", "R10CANCRE", "R11CANCRE", "R12CANCRE", "R13CANCRE")

### Cancer
names(dat |> select(contains("LUNGE"))) 
list_lung_4 <- c("R8LUNGE", "R10LUNGE", "R11LUNGE", "R12LUNGE","R13LUNGE")

### Heart problem
names(dat |> select(contains("HEARTE")))       
list_heart_5 <- c("R8HEARTE", "R10HEARTE",     
  "R11HEARTE", "R12HEARTE", "R13HEARTE")         
                                                 
### Stroke                                         
names(dat |> select(contains("STROKE")))
list_stroke_6 <- c("R8STROKE", "R10STROKE",
  "R11STROKE", "R12STROKE", "R13STROKE")

### Psychiatric Problems
names(dat |> select(contains("PSYCHE")))
list_psych_7 <- c("R8PSYCHE", "R10PSYCHE",
  "R11PSYCHE", "R12PSYCHE", "R13PSYCHE")

### Arthritis
names(dat |> select(contains("ARTHRE")))
list_arthr_8 <- c("R8ARTHRE", "R10ARTHRE",
  "R11ARTHRE", "R12ARTHRE", "R13ARTHRE")


## list Comorbidity index
 list_Comorbidity_index <- c(list_BP_1,         
  list_db_2, list_cancr_3, list_lung_4,          
  list_heart_5, list_stroke_6, list_psych_7,     
  list_arthr_8)                                  
                   

# try
#dat |> select(contains("H8ITOT")) |> head()
#dat |> select(contains("RAEDUC")) |> unique()
#dat |> select(list_Comorbidity_index) 

```



## data merge

```{r}
dat_an <- dat |> 
    select(list_id_husplb, list_IMRC, list_DLRC, list_SER7, list_confound, list_Comorbidity_index, list_exclude)
head(dat_an)
names(dat_an)
dim(dat_an)
```


## Make conbined variables

Outcome
- Z scored total recall
    - total recall = immediate word recall + delayed word recall(range from 0-20)
    - z scored total recall
    - mean and sd are used in the wave
- Z scored Series 7s
    - mean and sd are used in the wave

Confounders
- race and ethnicity 
    - Non-Hispanic White
    - Non-Hispanic Black
    - Hispanic
    - Non-Hispanic Other
- Comorbidity index
    - range from 0-8
    - total counts of Yes(1) among Doctor diagnosied health problems

```{r}

# Total recall, Z-scored total recall, Z-scored Series 7s
dat_an <- dat_an |>                    
      mutate(                            
          # Total recall = IMRC + DLRC   
          R8TR  = R8IMRC  + R8DLRC,     
          R10TR = R10IMRC + R10DLRC,
          R11TR = R11IMRC + R11DLRC,     
          R12TR = R12IMRC + R12DLRC,     
          R13TR = R13IMRC + R13DLRC,     
        # Z-scored total recall (within wave)                                 
          R8TR_Z=(R8TR  - mean(R8TR, na.rm = TRUE)) / sd(R8TR,na.rm =TRUE),
          R10TR_Z=(R10TR - mean(R10TR, na.rm = TRUE)) / sd(R10TR, na.rm = TRUE),
          R11TR_Z=(R11TR - mean(R11TR, na.rm = TRUE)) / sd(R11TR, na.rm = TRUE),
          R12TR_Z=(R12TR - mean(R12TR, na.rm = TRUE)) / sd(R12TR, na.rm = TRUE),
          R13TR_Z=(R13TR - mean(R13TR, na.rm = TRUE)) / sd(R13TR, na.rm = TRUE),
        # Z-scored SER7 (within wave)
          R8SER7_Z = (R8SER7 - mean(R8SER7, na.rm = TRUE)) / sd(R8SER7,
            na.rm = TRUE),
          R10SER7_Z = (R10SER7 - mean(R10SER7, na.rm = TRUE)) / sd(R10SER7, 
            na.rm = TRUE),
          R11SER7_Z = (R11SER7 - mean(R11SER7, na.rm = TRUE)) / sd(R11SER7, 
            na.rm = TRUE),
          R12SER7_Z = (R12SER7 - mean(R12SER7, na.rm = TRUE)) / sd(R12SER7,
            na.rm = TRUE),
          R13SER7_Z = (R13SER7 - mean(R13SER7, na.rm = TRUE)) / sd(R13SER7, 
            na.rm = TRUE)
      )


# Race and Ethnicity
 dat_an <- dat_an |>                 
      mutate(                       
          Race_etn = case_when(       
              RAHISPAN == 1 ~ "Hispanic",                         
              RARACEM == 1  ~ "Non-Hispanic White",               
              RARACEM == 2  ~ "Non-Hispanic Black",               
              RARACEM == 3  ~ "Non-Hispanic Other"
          )
      )

# Comorbidity index
 dat_an <- dat_an |>                 
      mutate(                         
          comob_ind_w8  = R8HIBPE  + R8DIABE + R8CANCRE  + R8LUNGE + R8HEARTE  + R8STROKE  + R8PSYCHE  + R8ARTHRE,
          comob_ind_w10 = R10HIBPE + R10DIABE + R10CANCRE + R10LUNGE + R10HEARTE + R10STROKE + R10PSYCHE + R10ARTHRE,                          
          comob_ind_w11 = R11HIBPE + R11DIABE + R11CANCRE + R11LUNGE +
            R11HEARTE + R11STROKE + R11PSYCHE + R11ARTHRE,
          comob_ind_w12 = R12HIBPE + R12DIABE + R12CANCRE + R12LUNGE +
            R12HEARTE + R12STROKE + R12PSYCHE + R12ARTHRE,
          comob_ind_w13 = R13HIBPE + R13DIABE + R13CANCRE + R13LUNGE +
            R13HEARTE + R13STROKE + R13PSYCHE + R13ARTHRE
      )

head(dat_an)
names(dat_an)
dim(dat_an)
```


## Exclusion

- Participants without exposure and outcomes data at baseline (Wave 8)  
- Participants who reported a doctor-diagnosed memory-related disease at baseline (Wave 8)  


```{r}

```



## Change into long format

```{r}
# check excluded number
nrow(dat_an)  # 45234 (Before exclusion)     
# (26,765 excluded)          
nrow(dat_an |> filter(!is.na(R8MEMRYE))) # 18469 remain
# (978 excluded)
nrow(dat_an |> filter(R8MEMRYE == 0)) # 17491 remain 
# (10,304 excluded)
nrow(dat_an |> filter(R8MEMRYE == 0, !is.na(R8LBHOUSEPRB))) # 7187 remain
# (90 excluded)
nrow(dat_an |> filter(R8MEMRYE == 0, !is.na(R8LBHOUSEPRB),
                    !is.na(R8IMRC))) # remain 7097
nrow(dat_an |> filter(R8MEMRYE == 0, !is.na(R8LBHOUSEPRB),
                    !is.na(R8IMRC), !is.na(R8DLRC))) # remain 7097
nrow(dat_an |> filter(R8MEMRYE == 0, !is.na(R8LBHOUSEPRB),
                    !is.na(R8IMRC), !is.na(R8DLRC),!is.na(R8SER7) ))
                    # remain 7097


dat_an_ex <- dat_an |> 
    filter(R8MEMRYE == 0, !is.na(R8LBHOUSEPRB),
            !is.na(R8IMRC), !is.na(R8DLRC),!is.na(R8SER7))

dim(dat_an_ex)

table(dat_an$R8LBHOUSEPRB, useNA = "always")
# Wave 8でhouseing stressがある人のwave 10, 11, 12, 13での状況
dat_an |> 
  filter(!is.na(R8LBHOUSEPRB)) |> 
  summarise(
    w10_housing = sum(!is.na(R10LBHOUSEPRB)),
    w11_housing = sum(!is.na(R11LBHOUSEPRB)),
    w12_housing = sum(!is.na(R12LBHOUSEPRB)),
    w13_housing = sum(!is.na(R13LBHOUSEPRB))
  ) |> view()

dat_an |> 
  filter(!is.na(R8LBHOUSEPRB)) |> 
  summarise(
    w14_housing = sum(!is.na(R14LBHOUSEPRB))
  )


# check num of w8 parricipants
## 18469
## No data in R8MEMRYE -> No data other variables
sum(!is.na(dat_an$R8MEMRYE)) # 18469 : w8 participants
sum(is.na(dat_an$R8MEMRYE)) # 26765: did not participante in w8
sum(is.na(dat_an$R8MEMRYE) & !is.na(dat_an$R8LBHOUSEPRB)) # 0
sum(is.na(dat_an$R8MEMRYE) & !is.na(dat_an$R8IMRC)) # 0
sum(is.na(dat_an$R8MEMRYE) & !is.na(dat_an$R8DLRC)) # 0
sum(is.na(dat_an$R8MEMRYE) & !is.na(dat_an$R8SER7)) # 0




```




## supplemental analysis
```{r}

view(dat |> filter(dat$HHIDPN == 10001010) |> select(contains("DLRC")))

dat_ex_R <- dat |> 
    select("HHIDPN", "R8LBHOUSEPRB", "R8DLRC", "R10DLRC", "R11DLRC", 
    "R12DLRC", "R13DLRC") 
head(dat_ex)
dim(dat_ex)

dat_ex |> 
    group_by(R8LBHOUSEPRB) |> 
    summarise(
        mean_delayed_recall = mean(R8DLRC, na.rm = TRUE),
    n = n()
    )

dat_ex |> 
    group_by(S8LBHOUSEPRB) |> 
    summarise(
        mean_delayed_recall = mean(S8DLRC, na.rm = TRUE),
        n = n()
    )

    dat_ex |> 
    group_by(S15LBHOUSEPRB) |> 
    summarise(
        mean_delayed_recall = mean(S15DLRCP, na.rm = TRUE),
        n = n()
    )

dat_ex_R <- dat_ex_R |> 
    drop_na("HHIDPN", "R8LBHOUSEPRB", "R8DLRC", "R10DLRC", "R11DLRC", 
    "R12DLRC", "R13DLRC")
dim(dat_ex_R)
dat_ex_R |> 
    group_by(R8LBHOUSEPRB) |> 
    summarise(
        mean_dlrc_w8 = mean(R8DLRC, na.rm = TRUE),
        mean_dlrc_w10 = mean(R10DLRC, na.rm = TRUE),

        mean_dlrc_w11 = mean(R11DLRC, na.rm = TRUE),
        mean_dlrc_w12 = mean(R12DLRC, na.rm = TRUE),
        mean_dlrc_w13 = mean(R13DLRC, na.rm = TRUE),
        n = n()
    )
dim(dat_ex_R)

library(ggplot2)
library(tidyr)

# データを作成して、long形式に変換
plot_data <- dat_ex_R |> 
    group_by(R8LBHOUSEPRB) |> 
    summarise(
        mean_dlrc_w8 = mean(R8DLRC, na.rm = TRUE),
        mean_dlrc_w10 = mean(R10DLRC, na.rm = TRUE),
        mean_dlrc_w11 = mean(R11DLRC, na.rm = TRUE),
        mean_dlrc_w12 = mean(R12DLRC, na.rm = TRUE),
        mean_dlrc_w13 = mean(R13DLRC, na.rm = TRUE),
        n = n()
    ) |>
    pivot_longer(
        cols = starts_with("mean_dlrc"),
        names_to = "wave",
        values_to = "mean_dlrc"
    ) |>
    mutate(
        wave_num = case_when(
            wave == "mean_dlrc_w8" ~ 8,
            wave == "mean_dlrc_w10" ~ 10,
            wave == "mean_dlrc_w11" ~ 11,
            wave == "mean_dlrc_w12" ~ 12,
            wave == "mean_dlrc_w13" ~ 13
        ),
        R8LBHOUSEPRB = factor(R8LBHOUSEPRB)
    )

# 折れ線グラフ作成
ggplot(plot_data, aes(x = wave_num, y = mean_dlrc, 
                       color = R8LBHOUSEPRB, 
                       group = R8LBHOUSEPRB)) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    labs(
        title = "Mean DLRC Score by Housing Problem Status",
        x = "Wave",
        y = "Mean DLRC Score",
        color = "Housing Problem"
    ) +
    scale_x_continuous(breaks = c(8, 10, 11, 12, 13)) +
    theme_minimal()
```

```{r}
dat_ex_R_ser7 <- dat |>
    select(HHIDPN, R8LBHOUSEPRB, R8SER7, R10SER7, R11SER7,
    R12SER7, R13SER7) |>
    drop_na()

table(dat_ex_R_ser7$R8LBHOUSEPRB, dat_ex_R_ser7$R8SER7)


dim(dat_ex_R_ser7)
dat_ex_R_ser7 |>
    group_by(R8LBHOUSEPRB) |>
    summarise(
        mean_ser7_w8 = mean(R8SER7, na.rm = TRUE),
        mean_ser7_w10 = mean(R10SER7, na.rm = TRUE),
        mean_ser7_w11 = mean(R11SER7, na.rm = TRUE),
        mean_ser7_w12 = mean(R12SER7, na.rm = TRUE),
        mean_ser7_w13 = mean(R13SER7, na.rm = TRUE),
        n = n()
    )

# データを作成して、long形式に変換
plot_data_ser7 <- dat_ex_R_ser7 |>
    group_by(R8LBHOUSEPRB) |>
    summarise(
        mean_ser7_w8 = mean(R8SER7, na.rm = TRUE),
        mean_ser7_w10 = mean(R10SER7, na.rm = TRUE),
        mean_ser7_w11 = mean(R11SER7, na.rm = TRUE),
        mean_ser7_w12 = mean(R12SER7, na.rm = TRUE),
        mean_ser7_w13 = mean(R13SER7, na.rm = TRUE),
        n = n()
    ) |>
    pivot_longer(
        cols = starts_with("mean_ser7"),
        names_to = "wave",
        values_to = "mean_ser7"
    ) |>
    mutate(
        wave_num = case_when(
            wave == "mean_ser7_w8" ~ 8,
            wave == "mean_ser7_w10" ~ 10,
            wave == "mean_ser7_w11" ~ 11,
            wave == "mean_ser7_w12" ~ 12,
            wave == "mean_ser7_w13" ~ 13
        ),
        R8LBHOUSEPRB = factor(R8LBHOUSEPRB)
    )

# 折れ線グラフ作成
ggplot(plot_data_ser7, aes(x = wave_num, y = mean_ser7,
                       color = R8LBHOUSEPRB,
                       group = R8LBHOUSEPRB)) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    labs(
        title = "Mean SER7 Score by Housing Problem Status",
        x = "Wave",
        y = "Mean SER7 Score",
        color = "Housing Problem"
    ) +
    scale_x_continuous(breaks = c(8, 10, 11, 12, 13)) +
    theme_minimal()
```

