---
title: "Analysis"
author: "Takashi Iida"
date: today
format: html
html-table-processing: none
---
## Overview

This document creates tables and figures for the manuscript.

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(haven)
```

## Load clean data 

- raw data: dat
- analysis data: dat_an

```{r}
# data import
# install.packages("haven")


getwd()

base_directory <- "/Users/iidatakashi/Documents/Git/HRS-E2-fellowship/data"

setwd(base_directory)

dat <- read_sas("randhrs1992_2022v1.sas7bdat")

```

## Overall data exploration

```{r}
#| label: overall-exploration
#| echo: true
#| message: false


head(dat) 
dim(dat)  # n. of observation and n. of variables
names(dat) # names of variables
```

## Variables names
Variable structure
- [Person/Unit][Wave][Topic...]
    - 1st character: who the variable refers to
        - R = respondent (reference person)
        - S = spouse/partner
        - H = household
    - 2nd character: wave identifier
        - 1,2,3,... = specific wave
        - sometimes A = “all waves” / not wave-specific (varies by item)

person id: HHIDPN (HHIDPN: HHold ID + Person Number /Num)


```{r}


dat %>% group_by("HHIDPN") |> select(contains("HOUSEPRB")) 
nrow(dat %>% group_by("HHIDPN"))

```



## Exposure exploration

- Exposure is Ongoing housing problems. 
- Identifier is HOUSEPRB in the colmun names. 
- Includde waves from 8 to 13 (10 years)

```{r}
#| label: exposure-exploration
#| echo: false

# numbers of waves containing HOUSEPRB
dim(dat %>% select(contains("HOUSEPRB"))) #  16

# number of participants for HOUSEPRB at wave 8-15
## wave 9 did not have HOUSEPRB
dat |>
  summarise(
    w8_housing = sum(!is.na(R8LBHOUSEPRB)), # 7422
    w10_housing = sum(!is.na(R10LBHOUSEPRB)), # 5319
    w11_housing = sum(!is.na(R11LBHOUSEPRB)), # 1 (no f/u subgrp )
    w12_housing = sum(!is.na(R12LBHOUSEPRB)), # 4450
    w13_housing = sum(!is.na(R13LBHOUSEPRB)), # 3
    w14_housing = sum(!is.na(R14LBHOUSEPRB)), # 2800
    w15_housing = sum(!is.na(R15LBHOUSEPRB)) # 2800
  )

# waves: S&R 8,10,11,12,13,14,15,16
names(dat %>% select(contains("HOUSEPRB"))) 

# select person_id,  House problem from w8 to w13
# make lists 
list_id_husplb <- c("HHIDPN", "R8LBHOUSEPRB", "R10LBHOUSEPRB","R11LBHOUSEPRB", "R12LBHOUSEPRB", "R13LBHOUSEPRB", "R14LBHOUSEPRB" )
# names(dat |> select(list_id_husplb, list_IMRC))

# Spouse lists
list_id_husplb_spo <- c("HHIDPN", "S8LBHOUSEPRB", "S10LBHOUSEPRB","S11LBHOUSEPRB", "S12LBHOUSEPRB", "S13LBHOUSEPRB", "S14LBHOUSEPRB" )
```


## Outcome-exploration

- Immediate Word Recall: IMRC
    - memo: RwFIMRC, RwFIMRCP, and RwFIMRCW indicate whether the value was imputed (=1) or not (=0). -> we will not use
- Delayed Word Recall
    - memo: RwFDLRC, RwFDLRCP, and RwFDLRCW indicate whether the value was imputed (=1) or not (=0). 
- Serial 7’s
- Wave 9 (2008) was excluded from the analysis as housing stress data were not collected in that wave

```{r}

#| label: outcome-exploration
#| echo: false

# Immediate word recall
names(dat |> select(contains("IMRC")))
list_IMRC <- c("R8IMRC","R10IMRC","R11IMRC", "R12IMRC", "R13IMRC")
list_IMRC_spo <- c("S8IMRC","S10IMRC","S11IMRC", "S12IMRC", "S13IMRC")

# Delayed word recall
names(dat |> select(contains("DLRC")))
list_DLRC <- c("R8DLRC","R10DLRC","R11DLRC", "R12DLRC", "R13DLRC")
list_DLRC_spo <- c("S8DLRC","S10DLRC","S11DLRC", "S12DLRC", "S13DLRC")

# Series 7s
names(dat |> select(contains("SER7")))
list_SER7 <- c("R8SER7","R10SER7","R11SER7", "R12SER7", "R13SER7")
list_SER7_spo <- c("S8SER7","S10SER7","S11SER7", "S12SER7", "S13SER7")

```



## Confounding exploration
variables to exlude 
- memory-related disease
    - whether or not the Respondent reports a doctor has ever told
them that they had this condition in this or any previous wave
    - 0: No
    - 1: Yes
- memo
    - dementia, alzheimer disease were measured from wave 10


variabels in table 1
- age
    - age at the begging of the interview in years
- sex
    - 1: Male
    - 2: Female
- Race
    - 1.White/Caucasian 
    - 2.Black/African American 
    - 3.Other
- Ethnicity
    - 0.Not Hispanic
    - 1.Hispanic
- education
    - 1. Lt High-school
    - 2. GED, 
    - 3. High-school graduate
    - 4. some college
    - 5. college and above
    - 1. Less than high school
    - 2. high school and general educational diplomas
    - 3. some College
    - 4. ollege and above
- income
    - total household income
- smoking
    - current smoking status
    - 0: No
    - 1: Yes
- Doctor diagnosied health problems: ever have condition
    - high BP
    - diabetes
    - cancer
    - lung disease
    - heart problem
    - stroke
    - psychiatic problem
    - arthritis
    - answer
        - 0: No
        - 1: Yes
    - memo
        - sleep disorder is measured from w13, so not included

Sex, Education were not included in adjusted model because of the feature of the model

```{r}
# exclusion variabels

## dementia ever
names(dat |> select(contains("MEMRYE")))
list_exclude <- c("R8MEMRYE")
list_exclude_spo <- c("S8MEMRYE")

dat |> select(list_exclude) |> filter(R8MEMRYE==1) |> nrow() # 978
dat |> select(list_exclude_spo) |> filter(S8MEMRYE==1) |> nrow() # 431




# variabels in table 1

## age at the begging of the interview in years
names(dat |> select(contains("AGEY"))) # R8AGEY_B
names(dat |> select(contains("AGEY"))) # S8AGEY_B

## Gender of respondents
## 1: Male
## 2: Female
names(dat |> select(contains("RAGENDER"))) # RAGENDER

## Race
## 1.White/Caucasian 
## 2.Black/African American 
## 3.Other
names(dat |> select(contains("RARACEM"))) # RARACEM


## Ethnicity
## 0.Not Hispanic
## 1.Hispanic
names(dat |> select(contains("RAHISPAN"))) # RAHISPAN

## Education
## 1. Lt High-school
## 2. GED, 
## 3. High-school graduate
## 4. some college
## 5. college and above
# change into
## 1. Less than high school
## 2. high school and general educational diplomas
## 3. some College
## 4. ollege and above
names(dat |> select(contains("RAEDUC"))) # RAEDUC


## Income:
## total household income
## cont : nominal dollars
## from w8 to w13
# "H8ITOT"  "H9ITOT"  "H10ITOT" "H11ITOT"
# "H12ITOT" "H13ITOT"
names(dat |> select(contains("ITOT"))) 

## Smoking
## 0: No
## 1: Yes
names(dat |> select(contains("SMOKEN"))) # R8SMOKEN, S8SMOKEN

list_confound <- c(
    "R8AGEY_B", "R10AGEY_B",   # age
    "R11AGEY_B", "R12AGEY_B", "R13AGEY_B",
    "RAGENDER", # gender
    "RARACEM",  # race
    "RAHISPAN", # ethnicity
    "RAEDUC", # education
    "H8ITOT",  "H10ITOT", # income
    "H11ITOT","H12ITOT", "H13ITOT",
    "R8SMOKEN", "R10SMOKEN", "R11SMOKEN", # smoking
    "R12SMOKEN", "R13SMOKEN"
    )

list_confound_spo <- c(
    "S8AGEY_B", "S10AGEY_B",   # age
    "S11AGEY_B", "S12AGEY_B", "S13AGEY_B",
    "RAGENDER", # gender (time-invariant, same variable)
    "RARACEM",  # race (time-invariant, same variable)
    "RAHISPAN", # ethnicity (time-invariant, same variable)
    "RAEDUC", # education (time-invariant, same variable)
    "H8ITOT",  "H10ITOT", # income (household-level, same variable)
    "H11ITOT","H12ITOT", "H13ITOT",
    "S8SMOKEN", "S10SMOKEN", "S11SMOKEN", # smoking
    "S12SMOKEN", "S13SMOKEN"
    )


## Doctor diagnosied health problems
### High BP
names(dat |> select(contains("HIBP")))
list_BP_1     <- c("R8HIBPE", "R10HIBPE","R11HIBPE", "R12HIBPE","R13HIBPE")
list_BP_1_spo <- c("S8HIBPE", "S10HIBPE","S11HIBPE", "S12HIBPE","S13HIBPE")

### Diabetes
names(dat |> select(contains("DIABE")))
list_db_2     <- c("R8DIABE", "R10DIABE", "R11DIABE", "R12DIABE", "R13DIABE")
list_db_2_spo <- c("S8DIABE", "S10DIABE", "S11DIABE", "S12DIABE", "S13DIABE")

### Cancer
names(dat |> select(contains("CANCRE")))
list_cancr_3     <- c("R8CANCRE", "R10CANCRE", "R11CANCRE", "R12CANCRE", "R13CANCRE")
list_cancr_3_spo <- c("S8CANCRE", "S10CANCRE", "S11CANCRE", "S12CANCRE", "S13CANCRE")

### Lung disease
names(dat |> select(contains("LUNGE")))
list_lung_4     <- c("R8LUNGE", "R10LUNGE", "R11LUNGE", "R12LUNGE","R13LUNGE")
list_lung_4_spo <- c("S8LUNGE", "S10LUNGE", "S11LUNGE", "S12LUNGE","S13LUNGE")

### Heart problem
names(dat |> select(contains("HEARTE")))
list_heart_5     <- c("R8HEARTE", "R10HEARTE", "R11HEARTE", "R12HEARTE", "R13HEARTE")
list_heart_5_spo <- c("S8HEARTE", "S10HEARTE", "S11HEARTE", "S12HEARTE", "S13HEARTE")

### Stroke
names(dat |> select(contains("STROKE")))
list_stroke_6     <- c("R8STROKE", "R10STROKE", "R11STROKE", "R12STROKE", "R13STROKE")
list_stroke_6_spo <- c("S8STROKE", "S10STROKE", "S11STROKE", "S12STROKE", "S13STROKE")

### Psychiatric Problems
names(dat |> select(contains("PSYCHE")))
list_psych_7     <- c("R8PSYCHE", "R10PSYCHE", "R11PSYCHE", "R12PSYCHE", "R13PSYCHE")
list_psych_7_spo <- c("S8PSYCHE", "S10PSYCHE", "S11PSYCHE", "S12PSYCHE", "S13PSYCHE")

### Arthritis
names(dat |> select(contains("ARTHRE")))
list_arthr_8     <- c("R8ARTHRE", "R10ARTHRE", "R11ARTHRE", "R12ARTHRE", "R13ARTHRE")
list_arthr_8_spo <- c("S8ARTHRE", "S10ARTHRE", "S11ARTHRE", "S12ARTHRE", "S13ARTHRE")


## list Comorbidity index
list_Comorbidity_index <- c(list_BP_1,
  list_db_2, list_cancr_3, list_lung_4,
  list_heart_5, list_stroke_6, list_psych_7,
  list_arthr_8)

list_Comorbidity_index_spo <- c(list_BP_1_spo,
  list_db_2_spo, list_cancr_3_spo, list_lung_4_spo,
  list_heart_5_spo, list_stroke_6_spo, list_psych_7_spo,
  list_arthr_8_spo)                                  
                   

# try
#dat |> select(contains("H8ITOT")) |> head()
#dat |> select(contains("RAEDUC")) |> unique()
#dat |> select(list_Comorbidity_index) 

```



## data merge

```{r}
# data for respondents
# identifiers for household
# H8HHID: household identifiers(with info about spliting or not in the wave)
names(dat |> select(contains("HHID")))
list_ident <- c("H8HHID", "H9HHID", "H10HHID", "H11HHID", "H12HHID", "H13HHID")

dat_an <- dat |>
    select(list_id_husplb, list_IMRC, list_DLRC, list_SER7, list_confound, list_Comorbidity_index, list_exclude,
    list_ident) # # household identifier
head(dat_an)
names(dat_an)
dim(dat_an)

# data for spouses
# identifiers for household
names(dat |> select(contains("HHID")))
list_ident_spo <- c("H8HHID", "H9HHID", "H10HHID", "H11HHID", "H12HHID", "H13HHID")

dat_an_spo <- dat |>
    select(list_id_husplb_spo, list_IMRC_spo, list_DLRC_spo, list_SER7_spo, list_confound_spo, list_Comorbidity_index_spo, list_exclude_spo,
    list_ident_spo) # household identifier

head(dat_an_spo)
names(dat_an_spo)
dim(dat_an_spo)

```


## Make combined variables

Outcome
- Z scored total recall
    - total recall = immediate word recall + delayed word recall(range from 0-20)
    - z scored total recall
    - mean and sd are used in the wave
- Z scored Series 7s
    - mean and sd are used in the wave

Confounders
- race and ethnicity 
    - Non-Hispanic White
    - Non-Hispanic Black
    - Hispanic
    - Non-Hispanic Other
- Comorbidity index
    - range from 0-8
    - total counts of Yes(1) among Doctor diagnosied health problems

```{r}

# Total recall, Z-scored total recall, Z-scored Series 7s
dat_an <- dat_an |>                    
      mutate(                            
          # Total recall = IMRC + DLRC   
          R8TR  = R8IMRC  + R8DLRC,     
          R10TR = R10IMRC + R10DLRC,
          R11TR = R11IMRC + R11DLRC,     
          R12TR = R12IMRC + R12DLRC,     
          R13TR = R13IMRC + R13DLRC,     
        # Z-scored total recall (within wave)                                 
          R8TR_Z=(R8TR  - mean(R8TR, na.rm = TRUE)) / sd(R8TR,na.rm =TRUE),
          R10TR_Z=(R10TR - mean(R10TR, na.rm = TRUE)) / sd(R10TR, na.rm = TRUE),
          R11TR_Z=(R11TR - mean(R11TR, na.rm = TRUE)) / sd(R11TR, na.rm = TRUE),
          R12TR_Z=(R12TR - mean(R12TR, na.rm = TRUE)) / sd(R12TR, na.rm = TRUE),
          R13TR_Z=(R13TR - mean(R13TR, na.rm = TRUE)) / sd(R13TR, na.rm = TRUE),
        # Z-scored SER7 (within wave)
          R8SER7_Z = (R8SER7 - mean(R8SER7, na.rm = TRUE)) / sd(R8SER7,
            na.rm = TRUE),
          R10SER7_Z = (R10SER7 - mean(R10SER7, na.rm = TRUE)) / sd(R10SER7, 
            na.rm = TRUE),
          R11SER7_Z = (R11SER7 - mean(R11SER7, na.rm = TRUE)) / sd(R11SER7, 
            na.rm = TRUE),
          R12SER7_Z = (R12SER7 - mean(R12SER7, na.rm = TRUE)) / sd(R12SER7,
            na.rm = TRUE),
          R13SER7_Z = (R13SER7 - mean(R13SER7, na.rm = TRUE)) / sd(R13SER7, 
            na.rm = TRUE)
      )


# Race and Ethnicity
 dat_an <- dat_an |>                 
      mutate(                       
          Race_etn = case_when(       
              RAHISPAN == 1 ~ "Hispanic",                         
              RARACEM == 1  ~ "Non-Hispanic White",               
              RARACEM == 2  ~ "Non-Hispanic Black",               
              RARACEM == 3  ~ "Non-Hispanic Other"
          ) |> 
        factor(levels = c("Non-Hispanic White", 
                            "Non-Hispanic Black",
                            "Hispanic",
                            "Non-Hispanic Other")) 
      )

# Comorbidity index
 dat_an <- dat_an |>                 
      mutate(                         
          comob_ind_w8  = R8HIBPE  + R8DIABE + R8CANCRE  + R8LUNGE + R8HEARTE  + R8STROKE  + R8PSYCHE  + R8ARTHRE,
          comob_ind_w10 = R10HIBPE + R10DIABE + R10CANCRE + R10LUNGE + R10HEARTE + R10STROKE + R10PSYCHE + R10ARTHRE,                          
          comob_ind_w11 = R11HIBPE + R11DIABE + R11CANCRE + R11LUNGE +
            R11HEARTE + R11STROKE + R11PSYCHE + R11ARTHRE,
          comob_ind_w12 = R12HIBPE + R12DIABE + R12CANCRE + R12LUNGE +
            R12HEARTE + R12STROKE + R12PSYCHE + R12ARTHRE,
          comob_ind_w13 = R13HIBPE + R13DIABE + R13CANCRE + R13LUNGE +
            R13HEARTE + R13STROKE + R13PSYCHE + R13ARTHRE
      )

head(dat_an)
names(dat_an)
dim(dat_an)
```


```{r}
# for spouses
# Total recall, Z-scored total recall, Z-scored Series 7s
dat_an_spo <- dat_an_spo |>
      mutate(
          # Total recall = IMRC + DLRC
          S8TR  = S8IMRC  + S8DLRC,
          S10TR = S10IMRC + S10DLRC,
          S11TR = S11IMRC + S11DLRC,
          S12TR = S12IMRC + S12DLRC,
          S13TR = S13IMRC + S13DLRC,
        # Z-scored total recall (within wave)
          S8TR_Z  = (S8TR  - mean(S8TR,  na.rm = TRUE)) / sd(S8TR,  na.rm = TRUE),
          S10TR_Z = (S10TR - mean(S10TR, na.rm = TRUE)) / sd(S10TR, na.rm = TRUE),
          S11TR_Z = (S11TR - mean(S11TR, na.rm = TRUE)) / sd(S11TR, na.rm = TRUE),
          S12TR_Z = (S12TR - mean(S12TR, na.rm = TRUE)) / sd(S12TR, na.rm = TRUE),
          S13TR_Z = (S13TR - mean(S13TR, na.rm = TRUE)) / sd(S13TR, na.rm = TRUE),
        # Z-scored SER7 (within wave)
          S8SER7_Z  = (S8SER7  - mean(S8SER7,  na.rm = TRUE)) / sd(S8SER7,  na.rm = TRUE),
          S10SER7_Z = (S10SER7 - mean(S10SER7, na.rm = TRUE)) / sd(S10SER7, na.rm = TRUE),
          S11SER7_Z = (S11SER7 - mean(S11SER7, na.rm = TRUE)) / sd(S11SER7, na.rm = TRUE),
          S12SER7_Z = (S12SER7 - mean(S12SER7, na.rm = TRUE)) / sd(S12SER7, na.rm = TRUE),
          S13SER7_Z = (S13SER7 - mean(S13SER7, na.rm = TRUE)) / sd(S13SER7, na.rm = TRUE)
      )


# Race and Ethnicity
dat_an_spo <- dat_an_spo |>
      mutate(
          Race_etn = case_when(
              RAHISPAN == 1 ~ "Hispanic",
              RARACEM == 1  ~ "Non-Hispanic White",
              RARACEM == 2  ~ "Non-Hispanic Black",
              RARACEM == 3  ~ "Non-Hispanic Other"
          ) |>
        factor(levels = c("Non-Hispanic White",
                            "Non-Hispanic Black",
                            "Hispanic",
                            "Non-Hispanic Other"))
      )

# Comorbidity index
dat_an_spo <- dat_an_spo |>
      mutate(
          comob_ind_w8  = S8HIBPE  + S8DIABE  + S8CANCRE  + S8LUNGE  + S8HEARTE  + S8STROKE  + S8PSYCHE  + S8ARTHRE,
          comob_ind_w10 = S10HIBPE + S10DIABE + S10CANCRE + S10LUNGE + S10HEARTE + S10STROKE + S10PSYCHE + S10ARTHRE,
          comob_ind_w11 = S11HIBPE + S11DIABE + S11CANCRE + S11LUNGE +
            S11HEARTE + S11STROKE + S11PSYCHE + S11ARTHRE,
          comob_ind_w12 = S12HIBPE + S12DIABE + S12CANCRE + S12LUNGE +
            S12HEARTE + S12STROKE + S12PSYCHE + S12ARTHRE,
          comob_ind_w13 = S13HIBPE + S13DIABE + S13CANCRE + S13LUNGE +
            S13HEARTE + S13STROKE + S13PSYCHE + S13ARTHRE
      )

head(dat_an_spo)
names(dat_an_spo)
dim(dat_an_spo)
```


## Combine

```{r}


# make dat_an_ex ----------------------------------------
dat_an_ex <- dat_an |>
  filter(
    R8MEMRYE == 0,          # no memory-related disease at baseline
    !is.na(R8LBHOUSEPRB),   # baseline exposure exist
    !is.na(R8IMRC),         # baseline outcome (episodic) exist
    !is.na(R8DLRC),
    !is.na(R8SER7)          # baseline outcome (working) exist
  )
# 45,234 → 17,491 → 7,187 → 7,097
dim(dat_an_ex)  # 7097 109

dat_an_spo_ex <- dat_an_spo |>
  filter(
    S8MEMRYE == 0,          # no memory-related disease at baseline
    !is.na(S8LBHOUSEPRB),   # baseline exposure exist
    !is.na(S8IMRC),         # baseline outcome (episodic) exist
    !is.na(S8DLRC),
    !is.na(S8SER7)          # baseline outcome (working) exist
  )
dim(dat_an_spo_ex)  # 4805  109 (but overlap exists)


# ----------------------------------------------------
# 1. give role (respondent or spouse)
# ------------------------------------------------------------
dat_resp <- dat_an_ex |>
  mutate(role = "respondent")

dat_spo <- dat_an_ex_spo |>
  rename_with(~ str_replace(.x, "^S(\\d)", "R\\1")) |>
  mutate(
    role = "spouse",
    R14LBHOUSEPRB = NA_real_ # no 
  )

# ------------------------------------------------------------
# 2. Stack and check
# ------------------------------------------------------------
dat_wide_combined <- bind_rows(dat_resp, dat_spo) |>
  arrange(HHIDPN, role) |>  
  distinct(HHIDPN, .keep_all = TRUE)

cat("Stacked rows:", nrow(dat_wide_combined), "\n")                    # 7,744
cat("Unique HHIDPN:", n_distinct(dat_wide_combined$HHIDPN), "\n")      # 7,744
n_distinct(dat_wide_combined$H8HHID) # 648
table(dat_wide_combined$role) # respondent: 7097, spouse: 647
              
stopifnot(nrow(dat_wide_combined) == n_distinct(dat_wide_combined$HHIDPN))


# NA check
# all respondents and spouse have valid household ID
dat_wide_combined |>
  group_by(role) |>
  summarise(
    n = n(),
    n_hhid_na    = sum(is.na(H8HHID)),
    n_hhid_valid = sum(!is.na(H8HHID))
  )

dat_wide_combined |> head()

# check HHIDPN overlap
sum(duplicated(dat_wide_combined$HHIDPN)) # 0

# check HHIDPN overlap
sum(duplicated(dat_wide_combined$H8HHID)) # 2725

```


## Exclusion

- Participants without exposure and outcomes data at baseline (Wave 8)  
- Participants who reported a doctor-diagnosed memory-related disease at baseline (Wave 8)  


```{r}
# ------------------------------------------------------------
# Step 0: all participants
# ------------------------------------------------------------

n_unique <- function(ids_r, ids_s) length(union(ids_r, ids_s))
all_r <- dat_an$HHIDPN
all_s <- dat_an_spo$HHIDPN
n_unique(all_r, all_s) # All unique participants: 45234

# ------------------------------------------------------------
# Step 1: Wave 8 participants
# ------------------------------------------------------------
r_w8 <- dat_an     |> filter(!is.na(R8MEMRYE)) |> pull(HHIDPN)
s_w8 <- dat_an_spo |> filter(!is.na(S8MEMRYE)) |> pull(HHIDPN)
n_unique(r_w8, s_w8) # Wave 8 unique: 18469
n_unique(all_r, all_s) - n_unique(r_w8, s_w8) # Excluded:26765

# ------------------------------------------------------------
# Step 2: memory disease exclude
# ------------------------------------------------------------
r_mem <- dat_an     |> filter(R8MEMRYE == 0) |> pull(HHIDPN)
s_mem <- dat_an_spo |> filter(S8MEMRYE == 0) |> pull(HHIDPN)
n_unique(r_mem, s_mem) # No memory disease unique: 17890
n_unique(r_w8, s_w8) - n_unique(r_mem, s_mem) # Excluded: 579

# ------------------------------------------------------------
# Step 3: housing stress data exist
# ------------------------------------------------------------
r_hse <- dat_an     |> filter(R8MEMRYE == 0, !is.na(R8LBHOUSEPRB)) |> pull(HHIDPN)
s_hse <- dat_an_spo |> filter(S8MEMRYE == 0, !is.na(S8LBHOUSEPRB)) |> pull(HHIDPN)
n_unique(r_hse, s_hse) # With housing stress unique: 7765
n_unique(r_mem, s_mem) - n_unique(r_hse, s_hse) # Excluded: 10125

# ------------------------------------------------------------
# Step 4: cognitive data exist（final analytic sample）
# ------------------------------------------------------------
r_fin <- dat_an     |> filter(R8MEMRYE == 0, !is.na(R8LBHOUSEPRB),
                               !is.na(R8IMRC), !is.na(R8DLRC), !is.na(R8SER7)) |> pull(HHIDPN)
s_fin <- dat_an_spo |> filter(S8MEMRYE == 0, !is.na(S8LBHOUSEPRB),
                               !is.na(S8IMRC), !is.na(S8DLRC), !is.na(S8SER7)) |> pull(HHIDPN)

# final analytic sample 
n_unique(r_fin, s_fin) # 7744

# Excluded
n_unique(r_hse, s_hse) - n_unique(r_fin, s_fin) # 21


# ------------------------------------------------------------
# check roles
# ------------------------------------------------------------

# Respondent only
length(setdiff(r_fin, s_fin)) # 2939

# Spouse only
length(setdiff(s_fin, r_fin)) # 647

# Both
length(intersect(r_fin, s_fin)) # 4158

# ------------------------------------------------------------
# check House problems follow-up for subgroup
# ------------------------------------------------------------

#  be careful about being overlap below
# respondents
dat_an |>
  filter(!is.na(R8LBHOUSEPRB)) |>
  summarise(
    w8_housing = sum(!is.na(R8LBHOUSEPRB)), # 7422
    w10_housing = sum(!is.na(R10LBHOUSEPRB)), # 5319
    w11_housing = sum(!is.na(R11LBHOUSEPRB)), # 1 (no f/u subgrp )
    w12_housing = sum(!is.na(R12LBHOUSEPRB)), # 4450
    w13_housing = sum(!is.na(R13LBHOUSEPRB)), # 3
    w14_housing = sum(!is.na(R14LBHOUSEPRB)) # 2800
  )

# spouses
dat_an_spo |>
  filter(!is.na(S8LBHOUSEPRB)) |>
  summarise(
    w8_housing  = sum(!is.na(S8LBHOUSEPRB)), # 5038
    w10_housing = sum(!is.na(S10LBHOUSEPRB)), # 3409
    w11_housing = sum(!is.na(S11LBHOUSEPRB)), # 0
    w12_housing = sum(!is.na(S12LBHOUSEPRB)), # 2632
    w13_housing = sum(!is.na(S13LBHOUSEPRB)), # 1
    w14_housing = sum(!is.na(S14LBHOUSEPRB)) # 1582
  ) 

```


## Change into long format

Use wave 8, 10, 12

```{r}
waves <- c(8, 10, 12)

dat_long_combined <- map_dfr(waves, function(w) {
  dat_wide_combined |> 
    transmute(
      HHIDPN,
      HHID  = get(paste0("H", w, "HHID")),  # HHID by wave
      role,                                  
      wave  = w,

      # Time-invariant
      sex            = RAGENDER,
      race           = RARACEM,
      hispanic       = RAHISPAN,
      education      = RAEDUC,
      race_etn       = Race_etn,

      # Exposure
      housing_stress = get(paste0("R", w, "LBHOUSEPRB")),

      # Outcomes
      imrc           = get(paste0("R", w, "IMRC")),
      dlrc           = get(paste0("R", w, "DLRC")),
      ser7           = get(paste0("R", w, "SER7")),
      total_recall   = get(paste0("R", w, "TR")),
      total_recall_z = get(paste0("R", w, "TR_Z")),
      ser7_z         = get(paste0("R", w, "SER7_Z")),

      # Confounders
      age               = get(paste0("R", w, "AGEY_B")),
      income            = get(paste0("H", w, "ITOT")),
      smoking           = get(paste0("R", w, "SMOKEN")),
      comorbidity_index = get(paste0("comob_ind_w", w))
    )
}) |> 
  arrange(HHIDPN, wave)

# check
dim(dat_long_combined)  # 23232    20
n_distinct(dat_long_combined$HHIDPN) # unique individuals: 7744
n_distinct(dat_long_combined$HHID) # Unique households: 5,206

```


# chekc num of housing stress between waves

```{r}
# check num of housing stress between waves
## -> too small num of partcipants changing from 1->2~4
## it might be difficult to chekc fixed effect in categorical housing stress
dat_long_combined %>%
  select(HHIDPN, wave, housing_stress) %>%
  pivot_wider(names_from = wave, values_from = housing_stress, 
              names_prefix = "w") %>%
  count(w8, w10, w12) %>%
  arrange(desc(n)) 
```


## Factor, Binary, Numeric

- Change education category

```{r}
# housing stress -> dichotomized
# 0 = No housing problem
# 1 = Yes, any housing problem (2+3+4)
dat_long_combined <- dat_long_combined |> 
  mutate(
    housing_stress_bin = ifelse(housing_stress == 1, 0, 1)
    )

# wave -> factor
dat_long_combined <- dat_long_combined |> 
  mutate(wave = factor(wave, levels = c(8, 10, 12)))

 # sex -> factor
dat_long_combined <- dat_long_combined |> 
  mutate(
    sex = factor(
      sex,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ))

# education -> change category + factor 
dat_long_combined <- dat_long_combined |>
  mutate(
    education = factor(
      case_when(
        education == 1        ~ 1,  # Lt high school
        education %in% c(2, 3) ~ 2,  # GED + high-school graduate を統合
        education == 4        ~ 3,  # some college
        education == 5        ~ 4   # college and above
      ),
      levels = c(1, 2, 3, 4),
      labels = c(
        "Less than high school",
        "High school and general educational diplomas",
        "Some college",
        "College and above"
      )
    )
  )

# comorbidity_index, ser7 -> numeric
dat_long_combined <- dat_long_combined |> 
  mutate(
    comorbidity_index = as.numeric(comorbidity_index),
    ser7 = as.numeric(ser7)
    )

# number of participants experienced Housing stress more than 1 wave
dat_long_combined |> 
  group_by(HHIDPN) |> 
  summarise(ever_stress = max(housing_stress_bin, na.rm = TRUE)) |> 
  summarise(n_ever_stress = sum(ever_stress == 1))

# check raw vs binary housing stress
# correclty converted
dat_long_combined |> 
  count(housing_stress, housing_stress_bin)


# check
head(dat_long_combined, 20) |> view()
names(dat_long_combined)
```


## Death records

- Need to define death and censoring

- using RADDATE(date of death) and IWEND(interview end date), categolized into:
  - IWEND exist: alive at the wave
  - RADDATE exist & RADDATE < IWEND at the wave : death at the wave
  

```{r}

# ── step 1: exact death date and interview date from dat─────────
person_level <- dat |>
  select(HHIDPN, RADDATE,
         R8IWEND,  R10IWEND,  R12IWEND, # respondent
         S8IWEND,  S10IWEND,  S12IWEND) |> # spouse
  distinct() |>
  filter(HHIDPN %in% dat_long_combined$HHIDPN)


# ── step 2: combine cols regardless of roles─────────────────
role_info <- dat_long_combined |>
  distinct(HHIDPN, role)

person_level <- person_level |>
  left_join(role_info, by = "HHIDPN") |>
  mutate(
    IWEND_w8  = if_else(role == "respondent", R8IWEND,  S8IWEND),
    IWEND_w10 = if_else(role == "respondent", R10IWEND, S10IWEND),
    IWEND_w12 = if_else(role == "respondent", R12IWEND, S12IWEND)
  ) |>
  mutate(
    across(c(RADDATE, IWEND_w8, IWEND_w10, IWEND_w12),
           ~ as.Date(., origin = "1960-01-01"))
  ) |>
  mutate(
    has_death_record = !is.na(RADDATE)
  )


# ── Step 3: define wave base date──────────────────────────

# check num of NA for interview dates at waves
person_level |>
  summarise(
    na_IWEND_w8  = sum(is.na(IWEND_w8)),
    na_IWEND_w10 = sum(is.na(IWEND_w10)),
    na_IWEND_w12 = sum(is.na(IWEND_w12))
  )

# NA in interview dates -> median
wave_ref <- person_level |>
  mutate(
    ref_w8  = coalesce(IWEND_w8,  as.Date("2006-09-01")),
    ref_w10 = coalesce(IWEND_w10, as.Date("2010-09-01")),
    ref_w12 = coalesce(IWEND_w12, as.Date("2014-09-01"))
  )


# ── step 4:  status(alive / dead / dropout) at each wave────────
status <- wave_ref |>
  mutate(
    # Wave 8
    status_w8 = case_when(
      has_death_record & RADDATE < ref_w8 ~ "dead",
      !is.na(IWEND_w8)                    ~ "alive",
      TRUE                                ~ "censored"
    ),
    # Wave 10
    status_w10 = case_when(
      status_w8 == "dead"                  ~ "dead",
      has_death_record & RADDATE < ref_w10 ~ "dead",
      !is.na(IWEND_w10)                    ~ "alive",
      TRUE                                 ~ "censored"
    ),
    # Wave 12
    status_w12 = case_when(
      status_w10 == "dead"                 ~ "dead",
      has_death_record & RADDATE < ref_w12 ~ "dead",
      !is.na(IWEND_w12)                    ~ "alive",
      TRUE                                 ~ "censored"
    )
  ) |>
  select(HHIDPN, has_death_record,
         status_w8, status_w10, status_w12)


# ── Step 5: check────────────────────
status |> count(status_w8) # alive      7744
status |> count(status_w10) # dead: 776, censored: 457
status |> count(status_w12) # dead: 1684, censored: 669

# check: no one dead at wave 8: 0
status |> filter(status_w8 == "dead") #0

# check: alive -> alive（dead -> alive=0）
status |>
  filter(status_w10 == "alive" & status_w8 == "dead") #0


# -- Step 6: merge into dat_long_combined----------------
dat_long_combined <- dat_long_combined |>
  left_join(
    status |>
      pivot_longer(
        cols      = starts_with("status_w"),
        names_to  = "wave_label",
        values_to = "status"
      ) |>
      mutate(wave = factor(str_extract(wave_label, "\\d+"),
                           levels = c("8", "10", "12"))) |> 
      select(-wave_label),
    by = c("HHIDPN", "wave")
  )

# check status count
dat_long_combined |>
  group_by(wave, status) |>
  count()

# check status NA
dat_long_combined |>
  summarise(na_status = sum(is.na(status))) # 0

dat_long_combined |> dim() # 23232    21
dat_long_combined$HHIDPN |> n_distinct() # 7744


# step 7: merge into wide format--------------------------------

dat_wide_combined <- dat_wide_combined |>
  left_join(
    status |> select(HHIDPN, status_w8, status_w10, status_w12),
    by = "HHIDPN"
  )

# check
nrow(dat_wide_combined)  # 7744

dat_wide_combined |>
  summarise(
    na_status_w8  = sum(is.na(status_w8)), # 0
    na_status_w10 = sum(is.na(status_w10)), # 0
    na_status_w12 = sum(is.na(status_w12)) # 0
  )  

```


## Save data

```{r}
getwd()

saveRDS(dat_wide_combined, "dat_wide_combined.rds")
saveRDS(dat_long_combined,  "dat_long_combined.rds")
```